---
title: "psych-sci-figs"
format: html
params:
  update_data: true
  fn: "ex01-figs-in-psych-sci.csv"
---

## About

This page extracts information about the data visualizations we explored in [Exercise-01](../exercises/ex01-figs-in-psych-sci.qmd) from the shared Google Sheet.

## Survey

Direct link: <https://docs.google.com/spreadsheets/d/1X-3kWpGAACdkAJ8-AeqYiu5TS6tRzpfdgXBYyPooCPo/edit?gid=0#gid=0>

<iframe src="https://docs.google.com/spreadsheets/d/1X-3kWpGAACdkAJ8-AeqYiu5TS6tRzpfdgXBYyPooCPo/edit?gid=0#gid=0" width="640" height="800" frameborder="0" marginheight="0" marginwidth="0">Loading…</iframe>

## Preparation

First, we load the external packages (groups of R commands) that we will be using.

```{r}
#| warning: false
library('ggplot2')
library('dplyr')

r_functions <- list.files(file.path(here::here(), "src", "R"), "\\.R$", full.names = TRUE)

purrr::map(r_functions, source)
```

## Gathering

Next, we download the data from the Google Sheet where it is collected.
Dr. Gilmore has stored his Google account credentials in a special environment file that can be accessed by the R command `Sys.getenv("GMAIL_SURVEY")`.

::: {.callout-tip}
It's vital to be *very* careful when creating and sharing code like this that involves sensitive information like login credentials.

Gilmore likes to put credentials in an `.Renviron` file that lives in his home directory. 
This is a [recommended practice](https://support.posit.co/hc/en-us/articles/360047157094-Managing-R-with-Rprofile-Renviron-Rprofile-site-Renviron-site-rsession-conf-and-repos-conf).
On Mac OS and Linux, that's `~/.Renviron`. 
You can use the `usethis::edit_r_profile()` command *at the R console* (not the Terminal) to open your own `.Renviron` file.
In Gilmore's case, he has added the following line to that file:

```
GMAIL_SURVEY="<my-google-account>"
```

Here, he has substituted his Google account with credentials/access to the required files for `<my-google-account>`.
Then, when the R code below calls `Sys.getenv("GMAIL_SURVEY")`, the value of those credentials is returned as a text string.

Make sure to close and save the `.Renviron` file and restart your R session before testing this yourself.

:::

```{r, eval=params$update_data}
#| label: gather-data-from-google
#| message: false
#| echo: true
if (!dir.exists('csv')) {
  message("Creating missing `csv/`.")
  dir.create("csv")
}

if (params$update_data) {
  options(gargle_oauth_email = Sys.getenv("GMAIL_SURVEY"))
  googledrive::drive_auth()

  googledrive::drive_download(
    "PSYCH-490.003-Spr-2025-Psych-Viz",
    path = file.path("csv", params$fn),
    type = "csv",
    overwrite = TRUE
  )
  message("Data updated.")
} else {
  message("Using stored data.")
}
```

The data file has been saved as a comma-separated value (CSV) format data file in a special directory called `csv/`.

::: {.callout-note}
Because these data *might* contain sensitive or identifiable information, we only keep a local copy and do not share it publicly via GitHub.
This is achieved by adding the name of the data directory to a special `.gitignore` file.
:::

## Cleaning

Next we load the saved data file, and then proceed to clean it.

```{r, eval=params$update_data}
#| label: load-saved-data
#| message: false
#| echo: true
ex01 <-
  readr::read_csv(file.path("csv", params$fn), show_col_types = FALSE)
```

There are `{r} dim(ex01)[1]` responses.

These are the column/variable names.

```{r}
# Google Forms puts the full question in the top row of the data file.
# We use the names() function to extract and print the original questions.
ex01_qs <- names(ex01)
ex01_qs
```

For simplicity, we visualize below only those with non-empty URLs to the specific figure.

## Source samples

```{r many-tables-lapply, results = "asis", warning=FALSE}
these_figs <- ex01 |>
  filter(!is.na(url_to_figure))

res <- invisible(lapply(1:dim(these_figs)[1], return_img_chunk, df = these_figs))
cat(unlist(res), sep = "\n")
```
