---
title: "Cleaning survey data for Ex06"
format: html
---

## About

This page documents the data cleaning undertaken so that students could use data collected earlier in the semester as part of [Exercise 03](../exercises/ex03-making-data.qmd) for [Exercise 06](../exercises/ex06-figs-w-r.qmd).

Similar steps were described in the slides used on [March 25, 2025](../schedule.qmd#tuesday-march-25)

## Setup

```{r}
library(dplyr)
```

## Gathering

```{r}
survey_auto <- readr::read_csv("../include/csv/PSYCH-490.003-Exercise-3.csv",
                               show_col_types = FALSE)
dim(survey_auto)
names(survey_auto)
```

## Cleaning

::: {.callout-caution}
We drop the "Timestamp", "What is the date of your favorite holiday?" and "Comments" variables to eliminate any possibility that respondents can be identified.
:::

```{r}
survey_cleaned <- survey_auto |>
  select(-c("Timestamp", "What is the date of your favorite holiday?", 
            "Comments"))
```

Next, we create a data frame for the data dictionary.

```{r}
long_qs <- names(survey_cleaned)

survey_dd <- data.frame(question = long_qs)
```

We rename the long variable names to shorter, more useful ones.

```{r}
survey_cleaned <- survey_cleaned |>
  rename(fav_flavor = "Favorite Icecream Flavor",
                best_pet = "Best Pet Type",
         math_confidence = "How confident are you with your math skills? [How confident are you in your math skills?]",
         creativity = "How confident are you with your math skills? [How creative are you?]",
         n_concerts = "How many concerts have you gone to?",
         credits_this_semester = "How many credits are you taking this semester?")
```

We add the shortened names to the data dictionary.

```{r}
survey_dd$variable <- names(survey_cleaned)
survey_dd$unique_values[1] <- unique(survey_cleaned[,1])
survey_dd$unique_values[2] <- unique(survey_cleaned[,2])
survey_dd$unique_values[3] <- unique(survey_cleaned[,3])
survey_dd$unique_values[4] <- unique(survey_cleaned[,4])
survey_dd$unique_values[5] <- unique(survey_cleaned[,5])
survey_dd$unique_values[6] <- unique(survey_cleaned[,6])
```

## Save

We save CSVs of the cleaned data and the data dictionary.

```{r}
readr::write_csv(survey_dd, "../exercises/csv/ex06-data-dictionary.csv")
readr::write_csv(survey_cleaned, "../exercises/csv/ex06-data.csv")
```

