---
title: "Gestalt in the wild"
format: html
params:
  update_data: true
  google_sheet_name: 'PSYCH 490.003-Gestalt-Examples'
  fn: gestalt_examples.csv
---

## About

This page extracts information about illustrations of Gesltalt principles that students found as part of their work on  [2025-02-13](../slides/wk05-2025-02-13-sensation-to-perception.qmd). Data are pulled  from the shared Google Sheet.

## Data

Direct link: <https://docs.google.com/spreadsheets/d/1G8U_IPMP0x17Sfl-FQfF37DN19bDQFm0GO_UmQQzv7Q/edit?usp=sharing>

<iframe src="https://docs.google.com/spreadsheets/d/1G8U_IPMP0x17Sfl-FQfF37DN19bDQFm0GO_UmQQzv7Q/edit?usp=sharing" width="640" height="800" frameborder="0" marginheight="0" marginwidth="0">Loading…</iframe>

## Preparation

First, we load the external packages (groups of R commands) that we will be using.

::: {.callout-important}

The code uses the `quietly()` function from the `purrr` package to suppress most of the feedback.

:::

```{r}
#| warning: false
library('ggplot2')
library('dplyr')

# r_functions <- list.files(file.path(here::here(), "src", "R"), "\\.R$", full.names = TRUE)
# 
# purrr::map(r_functions, source) |>
#   purrr::quietly()
```

## Gathering

Next, we download the data from the Google Sheet where it is collected.
Dr. Gilmore has stored his Google account credentials in a special environment file that can be accessed by the R command `Sys.getenv("GMAIL_SURVEY")`.

::: {.callout-tip}
It's vital to be *very* careful when creating and sharing code like this that involves sensitive information like login credentials.

Gilmore likes to put credentials in an `.Renviron` file that lives in his home directory. 
This is a [recommended practice](https://support.posit.co/hc/en-us/articles/360047157094-Managing-R-with-Rprofile-Renviron-Rprofile-site-Renviron-site-rsession-conf-and-repos-conf).
On Mac OS and Linux, that's `~/.Renviron`. 
You can use the `usethis::edit_r_profile()` command *at the R console* (not the Terminal) to open your own `.Renviron` file.
In Gilmore's case, he has added the following line to that file:

```
GMAIL_SURVEY="<my-google-account>"
```

Here, he has substituted his Google account with credentials/access to the required files for `<my-google-account>`.
Then, when the R code below calls `Sys.getenv("GMAIL_SURVEY")`, the value of those credentials is returned as a text string.

Make sure to close and save the `.Renviron` file and restart your R session before testing this yourself.

:::

```{r, eval=params$update_data}
#| label: gather-data-from-google
#| message: false
#| echo: true
if (!dir.exists('csv')) {
  message("Creating missing `csv/`.")
  dir.create("csv")
}

if (params$update_data) {
  options(gargle_oauth_email = Sys.getenv("GMAIL_SURVEY"))
  googledrive::drive_auth()

  googledrive::drive_download(
    params$google_sheet_name,
    path = file.path("csv", params$fn),
    type = "csv",
    overwrite = TRUE
  )
  message("Data updated.")
} else {
  message("Using stored data.")
}
```

The data file has been saved as a comma-separated value (CSV) format data file in a special directory called `csv/`.

## Cleaning

Next we load the saved data file, and then proceed to clean it.

```{r, eval=params$update_data}
#| label: load-saved-data
#| message: false
#| echo: true
gestalt <-
  readr::read_csv(file.path("csv", params$fn), show_col_types = FALSE)
```

There are `{r} dim(gestalt)[1]` responses.

These are the column/variable names.

```{r}
# Google Forms puts the full question in the top row of the data file.
# We use the names() function to extract and print the original questions.
gestalt_qs <- names(gestalt)
gestalt_qs
```

For simplicity, we visualize below only those with non-empty URLs to the specific figure.

## Summary data

```{r}
figs_w_urls <- gestalt |>
  filter(!is.na(url_to_img)) |>
  filter(stringr::str_detect(url_to_img, "[png|jpg|webp]$"))
```

There were *n*=`{r} length(unique(gestalt$identifier))` unique respondents.

Of the `{r} dim(gestalt)[1]` responses from these individuals or teams, *n*=`{r} dim(figs_w_urls)[1]` had URLs we could link to directly with non-data-visualization figures. 

## Make helper functions

```{r}
make_img_markdown_from_url <- function(img_url = "https://www.psychologicalscience.org/redesign/wp-content/uploads/2015/12/TCD_fig_1.jpg",
                                       lightbox = "{.lightbox}") {
  assertthat::is.string(img_url)
  
  component_1 <- paste0("![](")
  component_2 <- paste0(img_url, ")", lightbox)
  paste0(component_1, component_2)
}


return_img_chunk <- function(i, df) {
  assertthat::is.number(i)
  assertthat::assert_that(i > 0)
  assertthat::assert_that(is.data.frame(df))
  
  df_i <- df[i,]
  chunk_hdr_txt <- paste0("### Figure ", i, "\n\n")
  chunk_hdr <- knitr::knit_expand(text = chunk_hdr_txt)
  
  if (!is.na(df_i$url_to_img)) {
    img_markdown <- make_img_markdown_from_url(df_i$url_to_img)
  } else {
    img_markdown = "**No direct link to figure**"
  }
  body_txt <- paste0("\n\n| Analyst | Principle | \n|---|---|\n", 
  "| ", df_i$identifier, " | ", df_i$gestalt_principle, "|")
  knitr::knit_child(
    text = c(chunk_hdr, img_markdown, body_txt),
    envir = environment(),
    quiet = TRUE
  )  
}
```

## Figures found

```{r many-tables-lapply, results = "asis", warning=FALSE}
res <- invisible(lapply(1:dim(figs_w_urls)[1], return_img_chunk, df = figs_w_urls))
cat(unlist(res), sep = "\n")
```

## Sites found

```{r}
gestalt |>
  filter(!is.na(url_src)) |>
  filter(stringr::str_detect(url_src, "^https://")) |>
  dplyr::select(url_src) |>
#  dplyr::mutate(url_src = paste0("<", url_src, ">")) |>
  knitr::kable(format = 'html')
```


